---
title: "Week 2 Assignment - Clustering"
output: html_document
---

```{r,echo=TRUE, comment=FALSE,warning=FALSE,message=FALSE}
set.seed(1990)
library(dendextend)
library(NbClust)
library(mclust)
library(gmodels)

library(tidyverse)
```



```{r,eval=TRUE,echo=TRUE}
seg_data <- read.csv(file = "./data/SelfStatedData.csv",row.names=1)
head(seg_data)
```




```{r}
std_seg_data <- scale(seg_data)
dist <- dist(std_seg_data, method = "euclidean")
as.matrix(dist)[1:5,1:5]
```



```{r}
set.seed(1990)
clust <- hclust(dist, method = "ward.D2")
plot(clust)

```

# Cut the dendogram at the distance 6

```{r}

dend <- dist %>% hclust(method = "ward.D2") %>% as.dendrogram
cutree(dend, h = 6) %>%
  unique() %>%
  length()

```



# Consider the four-cluster solution. Which cluster cares the most about the attribute Cell Connectivity?

```{r}

plot(clust)
h_cluster <- cutree(clust, 4)
rect.hclust(clust, k=4, border="red")

```


```{r}
hclust_summary <- aggregate(std_seg_data, by=list(h_cluster), FUN=mean)
hclust_summary
```

# According to the majority rule, what is the best number of clusters for this dataset?

```{r,echo=TRUE}
set.seed(1990)
NbClust(data=std_seg_data, min.nc=2, max.nc=15, index="all", method="ward.D2")

```


# Consider the two-cluster solution. Use the R aggregate() function to compute the attribute preferences by cluster. Which are the two most important attributes for cluster 1?


```{r}

h_2_cluster <- cutree(clust, 2)

hclust_summary <- aggregate(std_seg_data, by=list(h_2_cluster), FUN=mean)
hclust_summary

```

# Use the R function kmeans() to run a three-cluster analysis. What are the sizes of the resulting three clusters?

```{r}
set.seed(1990)
k_3_cluster <-kmeans(std_seg_data, 3, iter.max=100,nstart=100)
k_3_cluster
```


# What is the most important attribute for cluster with the second largest size?


```{r}
aggregate(std_seg_data, by=list(k_3_cluster$cluster), FUN=length)

```



```{r}
cluster_k_3_summary <- aggregate(std_seg_data, by=list(k_3_cluster$cluster), FUN=mean)
cluster_k_3_summary

```

# According to the majority rule, what is the best number of clusters for this dataset?

```{r, warning=FALSE}
set.seed(1990)
NbClust(data=std_seg_data, min.nc=2, max.nc=15, index="all", method="kmeans")
```


# What is the best model suggested by this method?

```{r}
set.seed(1990)
lca_clust <- Mclust(std_seg_data,verbose = FALSE)
summary(lca_clust)
```


# How many components (i.e., clusters) does this analysis suggest?

```{r}
set.seed(1990)
lca_clust_eei <- Mclust(std_seg_data, verbose = FALSE, modelName = "EEI")
summary(lca_clust_eei)
```



```{r}
lca_eei_clusters <- lca_clust_eei$classification

lca_eei_clust_summary <- aggregate(std_seg_data, by=list(lca_eei_clusters),FUN=mean)
lca_eei_clust_summary

```

